High-Efficiency, Small-Footprint Couplers Between Arbitrary Nanophotonic Waveguide Modes

### 2012-01-04 presentations/ob1_wg_paper/paper.tex
Half done editing ob1 paper

\begin{document}
\title{High-Efficiency, Small-Footprint Couplers Between 
    Arbitrary Nanophotonic Waveguide Modes in Two Dimensions}
\author{Jesse Lu and Jelena Vu\v{c}kovi\'{c}}
\address{Stanford University, Stanford, California, USA.}
\email{jesselu@stanford.edu}

\maketitle
\begin{abstract}
We develop a design algorithm which can design 
    high efficiency ($\sim$95\%), 
    small-footprint (1-4 square vacuum wavelengths)
    couplers between arbitrary nanophotonic waveguide modes.
Such a capability is extremely useful since it fundamentally allows for the
    design of all linear optical devices.
Our ``objective-first'' method is
    computationally fast (15 minutes on a single-core personal computer), 
    requires no trial-and-error, and
    does not require guessing a good starting design.
\end{abstract}
\ocis{230.7370, 130.3990.}

@@

% Why is waveguide mode conversion important?
\subsection{The importance of waveguide mode conversion}
Optical mode conversion, 
    the efficient transfer of photons from one guided mode to another,
    is a fundamental requirement in nanophotonics.
In fact, the problem of converting between nanophotonic modes is essentially
    the function of \emph{all} linear nanophotonic devices.
For instance, efficient conversion between waveguides modes
    is useful for:
    \begin{enumerate}
        \item Coupling to and from optical fiber\cite{fibergrating}, 
        to communicate with the outside world.
    \item Coupling between various nanophotonic waveguides, 
        since different waveguides are best suited for different applications.
        For example, ridge waveguides seem ideal for 
            low-loss transport\cite{ridge},
            but other waveguides, 
            such as photonic crystal waveguides or slot waveguides,
            may be better suited for slow-light\cite{pcslow} 
            or nonlinear optical devices based on 
            localized field intensities\cite{slotfocus}.
    \item Coupling between different materials systems such as
       passive, active\cite{active}, and metallic\cite{metallic} devices. 
    \end{enumerate}
Although efficient waveguide coupling is essential in any nanophotonic system,
    practical methods to design couplers have not yet been developed.

% Why are current solutions insufficient?
\subsection{Approaches to designing waveguide couplers}
% Brute force
Of the design strategies currently available,
    brute-force parameter search is the most often employed 
    because of its sheer simplicity.
Although it may be suitable for tuning existing designs,
    the parameter space for most practical devices is simply too large
    for such a strategy to be tractable.

% Adiabatic mode conversion (large devices, symmetry breaking)
Adiabatic mode conversion strategies have been succesful
    for certain fiber-waveguide\cite{fwadia} and 
    waveguide-waveguide\cite{wwadia} couplers,
    even if the resulting devices often require large footprints.
However, adiabatic approaches cannot be used in many important cases 
    such as coupling in the out-of-plane direction, or
    coupling between modes of opposite symmetry.

% Local optimization
On the other hand, optimization methods based on local topological derivatives 
    seem very promising\cite{deriv}
    in that they are both much faster than brute-force methods and
    more adaptable than adiabatic strategies.
However, these methods still carry a significant computational burden 
    in that every updated design must be simulated at least once.
Also, there is a significant burden for the user from whom
    a good initial design is usually required.

% How does your algorithm solve this problem?
% \subsection{Advantages of objective-first design}
In contrast, our method
\begin{itemize}
    \item does not employ brute-force parameter searches,
    \item does not require a good initial design,
    \item is computationally fast (no simulations required),
    \item generates couplers between seemingly arbitrary waveguide modes, and
    \item can generate these couplers within a very small footprint. 
\end{itemize}
We outline the general design strategy,
    as well as how it is applied to coupler design, in the section below.

\section{Methods}
% What is objective-first optimization?
\subsection{Objective-first optimization}
The typical approach to designing physical structures can be formulated 
    in the following way, 
    where $x$ is the field variable and $p$ is the structure variable,
    \begin{subequations}\label{eq:adj}
    \begin{align} 
    \text{decrease} & \quad f(x) \label{eq:adj:obj} \\ 
    \text{subject to} & \quad g(x,p) = 0. \label{eq:adj:con}
    \end{align}
    \end{subequations}
Here, $f(x)$, the \emph{design objective}, 
    calculates the performance of the device 
    (e.g. amount of power not coupled to output mode); 
    while $g(x,p)$ is the underlying physical equation for the system
    (e.g. the electromagnetic wave equation).

In contrast, the objective-first formulation is
    \begin{subequations}\label{eq:ob1}
    \begin{align} 
    \text{decrease} & \quad \|g(x,p)\|^2 \label{eq:ob1:obj} \\ 
    \text{subject to} & \quad f(x) = 0, \label{eq:ob1:con}
    \end{align}
    \end{subequations}
    where $\|g(x,p)\|^2$ is the \emph{physics residual}.
We term this formulation ``objective-first''
    because the design objective is prioritized even above satisfying physics;
    specifically, we force our design to always exhibit the desired performance
    ($f(x) = 0$).

The differences between Eqs.~\ref{eq:adj} and \ref{eq:ob1} are:
\begin{enumerate}
    \item in Eq.~\ref{eq:adj:obj}, we attempt to decrease the design objective,
        while in Eq.~\ref{eq:ob1:con}, the design objective is kept at zero
        (i.e. we force the optimization to satisfy the design objective);
    \item in Eq.~\ref{eq:adj:con}, we always satisfy the underlying physics,
        while in Eq.~\ref{eq:ob1:obj}, physics is not necessarily satisfied,
            since the physics residual is generally non-zero
            (although the optimization is directed at minimizing this residual).
\end{enumerate}

Thus, the fundamental innovation in the objective-first approach
    is simply this:
    we forcibly impose the desired performance on the device at the expense of
    not perfectly satisfying the physical equation which governs its operation.

% What are the implications of the objective-first approach?
\subsection{Numerical implications of the objective-first approach}

While the differences between the design strategies presented in 
    Eqs.~\ref{eq:adj} and \ref{eq:ob1} are straightforward,
    the numerical implications are more subtle.

The first practical implication of the objective-first approach 
    is that the number of independent variables is increased to include
    both $x$ and $p$.
This occurs because in Eq.~\ref{eq:adj}, 
    the constraint that physics must be satisfied, $g(x,p)=0$, 
    essentially forces $x$ to be a dependent variable,
    since the choice of $p$ implicitly determines the value of $x$
    (there is generally a one-to-one mapping from $p$ to $x$).
In contrast, Eq.~\ref{eq:ob1} allows both $x$ and $p$ to vary independently,
    because the constraint, $f(x)=0$, is only a function of $x$.

Secondly, the amount of computation needed to enforce the constraint is
    drastically reduced in our objective-first approach.
This is because Eq.~\ref{eq:adj} requires a full solution of $g(x,p)=0$
    (i.e. a full simulation of the structure, $p$) to compute $x$.
In contrast, the constraint $f(x)=0$ in Eq.~\ref{eq:ob1} 
    can often be enforced with a trivial amount of computation (as shown below).

Lastly, the objective-first approach eliminates the need for even a reasonable
    initial design.
Generally, methods based on Eq.~\ref{eq:adj} require an initial design which
    already provides some limited functionality
    (e.g. a coupler which already transfers 
    a non-zero amount of power to the desired output mode).
In constrast, out method, based on Eq.~\ref{eq:ob1}, performs just as well
    when started from a completely non-functional design 
    (e.g. a coupler which transfers no power into the desired output mode).

% How is it applied to waveguide coupler design?
\subsection{Objective-first approach to waveguide coupler design}
We now apply the objective-first approach to the problem of designing 
    two-dimensional nanophotonic waveguide couplers.

We choose to work in the two-dimensional transverse electric mode,
    which only couples $E_x$, $E_y$, and $H_z$ ($E_z, H_x, H_y = 0$),
    since it is most relevant for on-chip devices.
We choose to use $H_z$ as the field variable ($x$ in Eq.~\ref{eq:ob1}), 
    and $\epsilon^{-1}$ (inverse of the permittivity) 
    as the structure variable $p$.
This results in the following representation of the physics residual, 
    based on the time-harmonic electromagnetic wave equation without sources;
    \begin{equation}
    \|g(H_z, \epsilon^{-1})\|^2 = 
    \| \nabla \times \epsilon^{-1} \nabla \times H_z - \mu_0 \omega^2 H_z \|^2,
    \end{equation}
    where $\omega$ is the angular frequency,
    and $\mu_0$ is the permeability of free-space.

@@

\begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{intro} 
    \caption{Boundary-value formulation of the design objective.
        The values of $H_z^\text{perfect}$, 
            defined along the dashed box surrounding the design area
            (coupling structure), 
            are shown in orange (color online).
        The values of $H_z^\text{perfect}$ along the top and bottom edges
            of the dashed box are set to zero.
        In this schematic, the fundamental and second-order waveguide modes
            have been chosen as the input and output modes respectively.}
    \label{fig:intro}
\end{figure}

@@

Such a design objective is both extremely simple and widely adaptable 
    to the design of nearly every kind of nanophotonic device.
Most importantly, it is trivial to enforce,
    requiring only that we overwrite boundary field values.
Although there is ambiguity in the relative phases of 
    the input and output boundary fields,
    our experience suggests that successful designs are possible for 
    arbitrary choice of relative phase.

@@

\subsection{Numerical methods used to solve the objective-first design problem}
The design problem is now
    \begin{subequations}\label{eq:act}
    \begin{align} 
    \text{decrease} & \quad  
        \| \nabla \times \epsilon^{-1} \nabla \times H_z - 
            \mu_0 \omega^2 H_z \|^2 \label{eq:act:obj} \\ 
    \text{subject to} & \quad 
        \begin{bmatrix}
        H_z - H_z^\text{perfect} \\
        \frac{\partial H_z}{\partial n} - 
            \frac{\partial H_z^\text{perfect}}{\partial n}
        \end{bmatrix}_\text{boundary}
        = 0.
    \end{align}
    \end{subequations}

@@

\section{Results}

@@

    \item coupling between waveguides of different refractive index and width,
    \item coupling between waveguide modes of different order and symmetry,

@@

        i.e., between a slab waveguide and a photonic crystal fiber. 

@@

        waveguide, and
    \item coupling from a dielectric waveguide to a (plasmonic) metal wire,

@@

    the results of which are shown in Figs.~\ref{fig:fiber}-\ref{fig:wire}.

@@

These results clearly demonstrate that our method
    \begin{itemize}
    \item can be applied to arbitrary input and output modes,
        as seen from the diverse selection of desired output waveguide modes;
    \item generates highly efficient devices with  
        typical coupling efficiencies of $95\%$;
    \item generates compact devices with footprints of only
        1-4 square vacuum wavelengths;
    \item does not require a good initial design.
        The initial design was simply $\epsilon = 9$ everywhere
            (a somewhat arbitrary guess, other values work as well),
            which in almost every case
            produces an initially non-functional device 
            (coupling efficiency of $0\%$);
    \item is computationally fast, 
        requiring only 15 minutes on a single-core personal computer.
    \end{itemize}

Critically, these results form a key advancement in our understanding of 
    nanophotonic design; namely, 
    that it is indeed possible to efficiently design
    virtually any linear optical device using an objective-first approach.

@@

We then apply an objective-first approach to the design of 
    five practical nanophotonic waveguide couplers which are difficult, 
    at best, to solve with existing methods. 
We show that our method produces
    high-efficiency designs ($\sim 95\%$ efficiency) 
    in small footprints ($\sim 1$ square vacuum wavelength)
    between arbitrary waveguide modes.
Furthermore, out method
    is computationally fast (15 minutes on a single-core personal computer), and
    does not require trial-and-error, or even 
    a good starting design.\\

This work has been supported in part by the 
    AFOSR MURI for Complex and Robust On-chip Nanophotonics 
    (Dr. Gernot Pomrenke), grant number FA9550-09-1-0704.
The Matlab code used is available online\cite{code}.

>>>>

\begin{document}
\title{High-Efficiency, Small-Footprint Couplers Between 
    Arbitrary Nanophotonic Waveguide Modes}
\author{Jesse Lu and Jelena Vu\v{c}kovi\'{c}}
\address{Stanford University, Stanford, California, USA.}
\email{jesselu@stanford.edu}

\maketitle
\begin{abstract}
We develop an algorithm for designing
    high efficiency ($\sim$95\%), 
    small-footprint (1-4 square vacuum wavelengths)
    couplers between arbitrary nanophotonic waveguide modes
    in two dimensions.
Our ``objective-first'' method is
    computationally fast (15 minutes on a single-core personal computer), 
    requires no trial-and-error, and
    does not require guessing a good starting design.
We demonstrate designs for various coupling problems which suggest 
    that our method allows for the design of any linear optical device.
\end{abstract}
\ocis{230.7370, 130.3990.}

@@

% Why is waveguide mode conversion important?
Optical mode conversion, 
    the efficient transfer of photons from one guided mode to another,
    is a fundamental requirement in nanophotonics.
For instance, efficient conversion between waveguides modes
    is essential for:
    \begin{itemize}
        \item Coupling to and from optical fiber\cite{fibergrating}, 
        to communicate with the outside world.
    \item Coupling between various nanophotonic waveguides, 
        since different waveguides are best suited for different applications.
        For example, ridge waveguides seem ideal for 
            low-loss transport\cite{ridge},
            but other waveguides, 
            such as photonic crystal waveguides or slot waveguides,
            may be better suited for slow-light\cite{pcslow} 
            or nonlinear optical devices based on 
            localized field intensities\cite{slotfocus}.
    \item Coupling between different materials systems such as
       passive, active\cite{active}, and metallic\cite{metallic} devices. 
    \end{itemize}
In fact, the problem of converting between nanophotonic modes is essentially
    the function of all linear nanophotonic devices, since
    any linear device is completely characterized by its input and output modes.

In this paper, we propose a design algorithm which, 
    we believe, solves this problem.
Namely, we outline a method capable of designing nanophotonic devices
    to efficiently transfer photons between 
    arbitrary input and output modes,
    and we demonstrate this method by designing various nanophotonic couplers
    between different waveguides.

Furthermore, we show that our method 
     does not employ brute-force parameter searches;
     does not require a good initial design;
     is computationally fast (no simulations required);
     and can generate high efficiency couplers within a very small footprint. 
% 
% 
% % Why are current solutions insufficient?
% % Brute force
% Of the design strategies currently available,
%     brute-force parameter search is the most often employed 
%     because of its sheer simplicity.
% Although it may be suitable for tuning existing designs,
%     the parameter space for most practical devices is simply too large
%     for such a strategy to be tractable.
% 
% % Adiabatic mode conversion (large devices, symmetry breaking)
% Adiabatic mode conversion strategies have been succesful
%     for certain fiber-waveguide\cite{fwadia} and 
%     waveguide-waveguide\cite{wwadia} couplers,
%     even if the resulting devices often require large footprints.
% However, adiabatic approaches cannot be used in many important cases 
%     such as coupling in the out-of-plane direction, or
%     coupling between modes of opposite symmetry.

% How does your algorithm solve this problem?
% \subsection{Advantages of objective-first design}
\section{Objective-first formulation}
% What is objective-first optimization?
The typical approach to designing physical structures can be formulated 
    in the following way:
    \begin{subequations}\label{eq:adj}
    \begin{align} 
    \text{decrease} & \quad f(x) \label{eq:adj:obj} \\ 
    \text{subject to} & \quad g(x,p) = 0, \label{eq:adj:con}
    \end{align}
    \end{subequations}
    where $x$ is the field variable and $p$ is the structure variable.
Here, $f(x)$, the \emph{design objective}, 
    calculates the performance of the device 
    (e.g. amount of power not coupled to output mode); 
    while $g(x,p)$ is the underlying physical equation for the system
    (e.g. the electromagnetic wave equation).

In contrast, the objective-first formulation is
    \begin{subequations}\label{eq:ob1}
    \begin{align} 
    \text{decrease} & \quad \|g(x,p)\|^2 \label{eq:ob1:obj} \\ 
    \text{subject to} & \quad f(x) = 0, \label{eq:ob1:con}
    \end{align}
    \end{subequations}
    where $\|g(x,p)\|^2$ is the \emph{physics residual}.
We term this formulation ``objective-first''
    because the design objective is prioritized even above satisfying physics;
    specifically, we force our design to always exhibit the desired performance
    ($f(x) = 0$), even at the expense of
    not perfectly satisfying the underlying physics which governs its operation.

The motivation behind this formulation is two-fold.
First, to arrive at a locally-optimal design rapidly 
    by allowing $x$ and $p$ to vary independently, 
    as opposed to Eq.~\ref{eq:adj}
    where the value of $x$ is completely dependent on $p$.
Second, to increase the likelihood of arriving at a high efficiency design
    by forcibly imposing $f(x) = 0$,
    and thereby circumventing any local optima consisting of
    low-performance devices.

% How is it applied to waveguide coupler design?
\section{Objective-first design of arbitrary waveguide couplers}
We demonstrate the objective-first approach
    by designing waveguide couplers in two-dimensions.
Specifically, we work in the two-dimensional transverse electric mode,
    and choose $H_z$ as the field variable $x$,
    and $\epsilon^{-1}$ (the inverse of the permittivity) 
    as the structure variable $p$.
This results in the following form of the physics residual,
    \begin{equation}
    \|g(x,p)\|^2 = 
    \|g(H_z, \epsilon^{-1})\|^2 = 
    \| \nabla \times \epsilon^{-1} \nabla \times H_z - \mu_0 \omega^2 H_z \|^2,
    \end{equation}
    where $\omega$ is the angular frequency,
    and $\mu_0$ is the permeability of free-space.

@@

\begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{intro} 
    \caption{Boundary-value formulation of the design objective.
        The values of $H_z^\text{perfect}$, 
            defined along the dashed box surrounding the design area
            (coupling structure), 
            are shown in orange. 
        The values of $H_z^\text{perfect}$ along the top and bottom edges
            of the dashed box are set to zero.
        In this schematic, the fundamental and second-order waveguide modes
            have been chosen as the input and output modes respectively.}
    \label{fig:intro}
\end{figure}

@@

Such a design objective is both extremely simple and widely adaptable 
    to the design of nearly every kind of nanophotonic device.
Most importantly, it is trivial to enforce,
    requiring only that we overwrite boundary field values.
Our experience suggests that successful designs are possible for 
    arbitrary choice of relative phase between the input and output fields.

@@

The design problem is now
    \begin{subequations}\label{eq:act}
    \begin{align} 
    \text{decrease} & \quad  
        \| \nabla \times \epsilon^{-1} \nabla \times H_z - 
            \mu_0 \omega^2 H_z \|^2 \label{eq:act:obj} \\ 
    \text{subject to} & \quad 
        \begin{bmatrix}
        H_z - H_z^\text{perfect} \\
        \frac{\partial H_z}{\partial n} - 
            \frac{\partial H_z^\text{perfect}}{\partial n}
        \end{bmatrix}_\text{boundary}
        = 0.
    \end{align}
    \end{subequations}

@@

\subsection{Coupler designs}

@@

    \item coupling between waveguides of different refractive index and width
        (Fig.~\ref{fig:fiber});
    \item coupling between waveguide modes of different order and symmetry
        (Fig.~\ref{fig:mode});

@@

        i.e., between a slab waveguide and a photonic crystal fiber
        (Fig.~\ref{fig:aircore})

@@

        waveguide (Fig.~\ref{fig:mim}); and
    \item coupling from a dielectric waveguide to a (plasmonic) metal wire
        (Fig.~\ref{fig:wire}).

@@

@@

These results clearly demonstrate that our method can
    be applied to arbitrary input and output modes,
    as seen from the diverse selection of desired output waveguide modes.
Moreover, it produces highly efficient devices with  
    typical coupling efficiencies of $95\%$,
    which are at the same time compact 
    (with footprints of only 1-4 square vacuum wavelengths).
In addition, the method does not require a good initial guess for the design.
In our case the initial design was simply $\epsilon = 9$ everywhere
        (a somewhat arbitrary guess, other values work as well).
Finally, the method is computationally fast, 
    requiring only 15 minutes on a single-core personal computer.

These results form a key advancement in our understanding of 
    nanophotonic design; namely, 
    that it is indeed possible to efficiently design
    virtually any linear optical device using an objective-first approach.

@@

We then apply an objective-first approach to the design of 
    five practical nanophotonic waveguide couplers which are difficult
    to solve with existing methods. 
We show that our method produces
    high-efficiency designs ($\sim 95\%$ efficiency) 
    in small footprints ($\sim 1$ square vacuum wavelength)
    between arbitrary waveguide modes.
Furthermore, out method
    is computationally fast (15 minutes on a single-core personal computer), and
    does not require trial-and-error, or even 
    a good starting design.\\

This work has been supported by the 
    AFOSR MURI for Complex and Robust On-chip Nanophotonics 
    (Dr. Gernot Pomrenke), grant number FA9550-09-1-0704.
The Matlab code used is available online\cite{code}.
